---
title: "Customize"
author: "David Gerbing"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{5. Customize: Custom Features}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include=FALSE}
knitr::opts_chunk$set(fig.width=4, fig.height=3.5)
```

```{r}
library("lessR")
d <- Read("Employee")
```


## Color Palates

The `getColors()` function can generate a variety of color palates. Use them to, for example, fill the bars in a histogram or bar chart with the  `fill` parameter.



### Qualitative Scales

The default is a color spectrum of 12 `hcl` colors presented in the order in which they are assigned to discrete levels of a categorical variable. For clarity, the `pal` or _palette_ parameter is set to `"hues"`. The colors are generated with the base R `hcl()` function. To maintain equal brightness/intensity across the colors, all colors are based on chroma set at `c=65` and luminance set at `l=55`. 

```{r}
getColors("hues")
```

Display the color spectrum of 12 `hcl` colors ordered by hue from 0 by intervals of 360/12 = 30 degrees.

```{r}
getColors(in_order=TRUE)
```

Display a color wheel of 36 ordered hues around the wheel.

```{r}
getColors(n=36, shape="wheel", border="off")
```


Obtain deep rich colors for HCL qualitative scale.

```{r}
BarChart(Dept, fill=getColors(c=90, l=45))
```



### Sequential Scales

__lessR__ provides pre-defined sequential color scales across the range of hues around the color wheel: `"reds"`, `"rusts"`, `"browns"`, `"olives"`, `"greens"`, `"emeralds"`, `"turqoises"`, `"aquas"`, `"blues"`, `"purples"`, `"biolets"`, `"magentas"`, and `"grays"`.


To view the scales, use the function `showPalettes()`. Here it is commented out by the `#` sign in the first column because it writes the output to a pdf.

```{r}
#showPalettes()
```  

Generate a bar chart with blue hues.

```{r}
BarChart(Dept, fill=getColors("blues"))
```

Generate an `hcl` blue sequence with `c=60` and vary `l`.

```{r}
BarChart(Dept, fill=getColors("blues", c=60, l=c(30,80)))
```


Vary chroma for a yellow `hcl` sequence.

```{r}
BarChart(Dept, fill=getColors("browns", c=c(20,90), l=60))
```

Stack the three time series, fill under each curve with a version of the sequential range `"reds"` with some transparency.

```{r, fig.width=6}
d <- Read("StockPrice", quiet=TRUE)
Plot(date, Price, by=Company, trans=0.4, stack=TRUE, area_fill="reds")
d <- Read("Employee", quiet=TRUE)
```


### Divergent Scales

Generate colors from rust to blue.

```{r}
BarChart(Dept, fill=getColors("rusts", "blues"))
```

Add a custom value of chroma, `c`, to make less saturated.
```{r}
BarChart(Dept, fill=getColors("rusts", "blues", c=45))
```


```{r}
BarChart(Dept, fill=getColors(c=90, l=45))
```

### Manual Specification of Colors

Individually specify colors.

```{r}
getColors(c("black", "blue", "red"))
```

Generate a custom sequential range of colors.

```{r}
getColors(pal="aliceblue", end_pal="blue")
```


## Visualization Themes

The `style()` function specifies broad color themes, as well as individual characteristics of a visualization. Changes are persistent across subsequent visualizations unless again modified.

```{r}
style("darkred")
BarChart(Dept)
```

Return to the default `colors` for the `theme` parameter by either explicitly specifying, or just go with no specification.

```{r}
style()
```

Individual characteristics of a visualization can also be specified. Here change the background color of the visualization window.

```{r}
style(window_fill="aliceblue")
BarChart(Dept)
```

Return to the default theme, `"colors"`.
```{r}
style()
```

View all modifiable individual characteristics with the `show` parameter set to `TRUE`.

```{r}
style(show=TRUE)
```



## Full Manual

Use the base R `help()` function to view the full manual for `getColors()` or `style()`. Simply enter a question mark followed by the name of the function.

```
?getColors
?style
```

