---
title: "Introduction to lessR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width=5.5, fig.height=3,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lessR)
```

The vignette examples of using **lessR** became so extensive that the maximum R package installation size was exceeded. Find a limited number of examples below. Find many more vignette examples at:

[lessR examples](https://web.pdx.edu/~gerbing/lessR/examples/)


## Read Data

[more examples of reading and writing data files](https://web.pdx.edu/~gerbing/lessR/examples/ReadWrite.html)

Many of the following examples analyze data in the _Employee_ data set, included with **lessR**. To read an internal **lessR** data set, just pass the name of the data set to the **lessR** function `Read()`. Read the _Employee_ data into the data frame _d_. For data sets other than those provided by **lessR**, enter the path name or URL between the quotes, or leave the quotes empty to browse for the data file on your computer system. See the `Read and Write` vignette for more details.

```{r read}
d <- Read("Employee")
```

_d_ is the default name of the data frame for the **lessR** data analysis functions. Explicitly access the data frame with the `data` parameter in the analysis functions. 

As an option, also read the table of variable labels. Create the table formatted as two columns. The first column is the variable name and the second column is the corresponding variable label. Not all variables need be entered into the table. The table can be a `csv` file or an Excel file.

Read the file of variable labels into the _l_ data frame, currently the only permitted name. The labels will be displayed on both the text and visualization output. Each displayed label is the variable name juxtaposed with the corresponding label, as shown in the following output.

```{r labels}
l <- rd("Employee_lbl")
l
```

## Bar Chart

[more examples of bar charts and pie charts](https://web.pdx.edu/~gerbing/lessR/examples/BarChart.html)

Consider the categorical variable Dept in the Employee data table.  Use `BarChart()` to tabulate and display the visualization of the number of employees in each department, here relying upon the default data frame (table) named _d_. Otherwise add the `data=` option for a data frame with another name.

```{r bcEx, fig.width=4, fig.height=3.5, fig.align='center', fig.cap="Bar chart of tablulated counts of employees in each department."}
BarChart(Dept)
```

Specify a single fill color with the `fill` parameter, the edge color of the bars with `color`. Set the transparency level with `transparency`. Against a lighter background, display the value for each bar with a darker color using the `labels_color` parameter. To specify a color, use color names, specify a color with either its `rgb()` or `hcl()` color space coordinates, or use the **lessR** custom color palette function `getColors()`.

```{r fig.width=4, fig.height=3.75, fig.align='center'}
BarChart(Dept, fill="darkred", color="black", transparency=.8,
         labels_color="black")
```

Use the `theme` parameter to change the entire color theme: "colors", "lightbronze", "dodgerblue", "slatered", "darkred", "gray", "gold", "darkgreen", "blue", "red", "rose", "green", "purple", "sienna", "brown", "orange", "white", and "light". In this example, changing the full theme accomplishes the same as changing the fill color. Turn off the displayed value on each bar with the parameter `labels` set to `off`. Specify a horizontal bar chart with base R parameter `horiz`.

```{r fig.width=4, fig.height=3.5, fig.align='center'}
BarChart(Dept, theme="gray", labels="off", horiz=TRUE)
```

## Histogram

[more examples of histograms](https://web.pdx.edu/~gerbing/lessR/examples/Histogram.html)

Consider the continuous variable _Salary_ in the Employee data table.  Use `Histogram()` to tabulate and display the number of employees in each department, here relying upon the default data frame (table) named _d_, so the `data=` parameter is not needed.

```{r hs, fig.width=4, fig.height=3.5, fig.align='center', fig.cap="Histogram of tablulated counts for the bins of Salary."}
Histogram(Salary)
```

By default, the `Histogram()` function provides a color theme according to the current, active theme. The function also provides the corresponding frequency distribution, summary statistics, the table that lists the count of each category, from which the histogram is constructed, as well as an outlier analysis based on Tukey's outlier detection rules for box plots.

Use the parameters `bin_start`, `bin_width`, and `bin_end` to customize the histogram.

Easy to change the color, either by changing the color theme with `style()`, or just change the fill color with `fill`. Can refer to standard R colors, as shown with **lessR** function `showColors()`, or implicitly invoke the **lessR** color palette generating function `getColors()`. Each 30 degrees of the color wheel is named, such as `"greens"`, `"rusts"`, etc, and implements a sequential color palette.

```{r binwidth, fig.width=4, fig.height=3.5, fig.align='center', fig.cap="Customized histogram."}
Histogram(Salary, bin_start=35000, bin_width=14000, fill="reds")
```

## Scatterplot

[more examples of scatter plots and related](https://web.pdx.edu/~gerbing/lessR/examples/Plot.html)

Specify an X and Y variable with the plot function to obtain a scatter plot. For two variables, both variables can be any combination of continuous or categorical. One variable can also be specified. A scatterplot of two categorical variables yields a bubble plot. Below is a scatterplot of two continuous variables. 

```{r sp, fig.width=4}
Plot(Years, Salary)
```

Enhance the default scatterplot with parameter `enhance`. The visualization includes the mean of each variable indicated by the respective line through the scatterplot, the 95% confidence ellipse, labeled outliers, least-squares regression line with 95% confidence interval, and the corresponding regression line with the outliers removed. 

```{r spEnhance, fig.width=4}
Plot(Years, Salary, enhance=TRUE)
```

The default plot for a single continuous variable includes not only the scatterplot, but also the superimposed violin plot and box plot, with outliers identified. Call this plot the VBS plot.

```{r x1, fig.height=3}
Plot(Salary)
```

Following is a scatterplot in the form of a bubble plot for two categorical variables.

```{r spBubble, fig.width=4}
Plot(JobSat, Gender)
```

## Means and Proportions

### Means
[more examples of t-tests and ANOVA](https://web.pdx.edu/~gerbing/lessR/examples/Means.html)

For the independent-groups t-test, specify the response variable to the left of the tilde, `~`, and the categorical variable with two groups, the grouping variable, to the right of the tilde.

```{r, fig.height=4.25, fig.width=5}
ttest(Salary ~ Gender)
```

Next, to analyze the operational efficiency of a weeping device, do the two way independent groups ANOVA analyzing the variable _breaks_ across levels of _tension_ and _wool_. Specify the second independent variable preceded by a `*` sign.

```{r fig.width=5}
ANOVA(breaks ~ tension * wool, data=warpbreaks)
```

For a one-way ANOVA, just include one independent variable. A randomized block design is also available.

### Proportions

[more examples of analyzing proportions](https://web.pdx.edu/~gerbing/lessR/examples/Proportions.html)

The analysis of proportions is of two primary types. 

- Focus on a single value of a categorical variable, termed a "success" when it occurs, for one or more samples of data. Analyze the resulting proportion of occurrence for a single sample, or for a test of _homogeneity_, compare proportions of successes across distinct data samples for a single variable.
- Compare the obtained proportions across the values of one or more categorical variables for a single sample. Applied to a single variable, the analysis is a _goodness-of-fit_. Or, evaluate a potential relationship between two categorical variables, a test of _independence_.

Here, just analyze the $\chi^2$ test of independence, which applies to two categorical variables. The first categorical variable listed in this example is the value of the parameter `variable`, the first parameter in the function definition, so does not need the parameter name. The second categorical variable listed must include the parameter name `by`. 

The question for the analysis is if the observed frequencies of _Jacket_ thickness and _Bike_ ownership sufficiently differ from the frequencies expected by the null hypothesis that we conclude the variables are related.

```{r}
d <- Read("Jackets")
Prop_test(Jacket, by=Bike)
```

## Regression Analysis

[more examples of regression and logistic regression](https://web.pdx.edu/~gerbing/lessR/examples/Regression.html)

The full output is extensive: Summary of the analysis, estimated model, fit indices, ANOVA, correlation matrix, collinearity analysis, best subset regression, residuals and influence statistics, and prediction intervals. The motivation is to provide virtually all of the information needed for a proper regression analysis.

```{r}
d <- Read("Employee", quiet=TRUE)
reg(Salary ~ Years + Pre)
```

As with several other **lessR** functions, save the output to an object with the name of your choosing, such as `r`, and then reference desired pieces of the output. View the names of those pieces from the manual, here obtained with `?reg`, or use the R names function, such as in the following example.

```{r}
r <- reg(Salary ~ Years + Pre)
names(r)
```

View any piece of output with the name of the output file, a dollar sign, and the specific name of that piece. Here, examine the fit indices.

```{r}
r$out_fit
```

These expressions could also be included in a markdown document that systematically reviews each desired piece of the output.

## Time Series and Forecasting

[more examples of run charts, time series charts, and forecasting](https://web.pdx.edu/~gerbing/lessR/examples/Time.html)

The time series plot, plotting the values of a variable cross time, is a special case of a scatterplot, potentially with the points of size 0 with adjacent points connected by a line segment. Indicate a time series by specifying the `x`-variable, the first variable listed, as a variable of type `Date`. Unlike Base R functions, `Plot()` automatically converts to `Date` data values as dates specified in a digital format, such as `18/8/2024` or related formats plus examples such as `2024 Q3` or `2024 Aug`. Otherwise, explicitly use the R function `as.Date()` to convert to this format before calling `Plot()` or pass the date format directly with the `ts_format` parameter.

`Plot()` implements time series forecasting based on trend and seasonality with either exponential smoothing or regression analysis, including the accompanying visualization. Time series parameters include:

- `ts_method`: Set at `"es"` for exponential smoothing, the default, or `"lm"` for linear model regression.
- `ts_unit`: The time unit, either as the natural occurring interval between dates in the data, the default, or aggregated to a wider time interval.
- `ts_ahead`: The number of time units to forecast into the future
- `ts_agg`: If aggregating the time unit, aggregate as the `"sum"`, the default, or as the `"mean"`. 
- `ts_PIlevel`: The confidence level of the prediction intervals, with 0.95 the default.
- `ts_format`: Provides a specific format for the date variable if not detected correctly by default. 
- `ts_seasons`: Set to `FALSE` to turn off seasonality in the estimated model.
- `ts_trend`: Set to `FALSE` to turn off trend in the estimated model.
- `ts_type`: Applies to exponential smoothing to specify additive or multiplicative seasonality, with additive the default.

In this `StockPrice` data file, the date conversion as already been done.


```{r}
d <- Read("StockPrice")
head(d)
```

We have the date as _Month_, and also have variables _Company_ and stock _Price_.

```{r}
d <- Read("StockPrice")
Plot(Month, Price, filter=(Company=="Apple"), area_fill="on")
```


With the `by` parameter, plot all three companies on the same panel.

```{r}
Plot(Month, Price, by=Company)
```

Here, aggregate the mean by time, from months to quarters.


```{r}
Plot(Month, Price, ts_unit="quarters", ts_agg="mean")
```

`Plot()` implements exponential smoothing or linear regression with seasonality forecasting with accompanying visualization. Parameters include `ts_ahead` for the number of `ts_units` to forecast into the future, and `ts_format` to provide a specific format for the date variable if not detected correctly by default. Parameter `ts_method` defaults to `es` for exponential smoothing, or set to `lm` for linear regression. Control aspects of the exponential smoothing estimation and prediction algorithms with parameters `ts_level` (alpha), `ts_trend` (beta), `ts_seasons` (gamma), `ts_type` for additive or multiplicative seasonality, and `ts_PIlevel` for the level of the prediction intervals.

To forecast Apple's stock price, focus here on the last several years of the data, beginning with Row 400 through Row 473, the last row of data for apple. In this example, forecast ahead 24 months.


```{r}
d <- d[400:473,]
Plot(Month, Price, ts_unit="months", ts_agg="mean", ts_ahead=24)
```


## Factor Analysis

[more examples of exploratory and confirmatory factor analysis](https://web.pdx.edu/~gerbing/lessR/examples/FactorAnalysis.html)

Access the lessR data set called `datMach4` for the analysis of 351 people to the Mach IV scale. Read the optional variable labels. Including the item contents as variable labels means that the output of the confirmatory factor analysis contains the item content grouped by factor.

```{r}
d <- Read("Mach4", quiet=TRUE)
l <- Read("Mach4_lbl", var_labels=TRUE)
```

Calculate the correlations and store in here in *mycor*, a data structure that contains the computed correlation matrix with the name R. Extract R from *mycor*.

```{r fig.width=3.5, fig.height=3.5}
mycor <- cr(m01:m20)
R <- mycor$R
```

The correlation matrix for analysis is named R. The item (observed variable) correlation matrix is the numerical input into the confirmatory factor analysis.

Here, do the default two-factor solution with `"promax"` rotation. The default correlation matrix is *mycor*. The abbreviation for `corEFA()` is `efa()`.

``` {r}
efa(R, n_factors=4)
```


The confirmatory factor analysis is of multiple-indicator measurement scales, that is, each item (observed variable) is assigned to only one factor. Solution method is centroid factor analysis.

Specify the measurement model for the analysis in Lavaan notation. Define four factors: Deceit, Trust, Cynicism, and Flattery.

```{r}
MeasModel <- 
" 
   Deceit =~ m07 + m06 + m10 + m09 
   Trust =~ m12 + m05 + m13 + m01 
   Cynicism =~ m11 + m16 + m04 
   Flattery =~ m15 + m02 
"
```


## Pivot Tables

[more examples of pivot tables](https://web.pdx.edu/~gerbing/lessR/examples/pivot.html)

Aggregate with `pivot()`. Any function that processes a single vector of data, such as a column of data values for a variable in a data frame, and outputs a single computed value, the statistic, can be passed to `pivot()`. Functions can be user-defined or built-in. 

Here, compute the mean and standard deviation of each company in the StockPrice data set download it with **lessR**.

```{r}
d <- Read("StockPrice", quiet=TRUE)
pivot(d, c(mean, sd), Price, by=Company)
```

Interpret this call to `pivot()` as

- for data set _d_
- compute the mean and standard deviation 
- of variable _Price_ 
- for each _Company_

Select any two of the three possibilities for multiple parameter values: Multiple compute functions, multiple variables over which to compute, and multiple categorical variables by which to define groups for aggregation.


## Color Scales

[more examples of color scales](https://web.pdx.edu/~gerbing/lessR/examples/Customize.html)


Generate color scales with `getColors()`. The default output of `getColors()` is a color spectrum of 12 `hcl` colors presented in the order in which they are assigned to discrete levels of a categorical variable. For clarity in the following function call, the default value of the `pal` or _palette_ parameter is explicitly set to its name, `"hues"`.


```{r}
getColors("hues")
```

```{r echo=FALSE}
getColors("hues", output=TRUE)
```



__lessR__ provides pre-defined sequential color scales across the range of hues around the color wheel in 30 degree increments: `"reds"`, `"rusts"`, `"browns"`, `"olives"`, `"greens"`, `"emeralds"`, `"turqoises"`, `"aquas"`, `"blues"`, `"purples"`, `"biolets"`, `"magentas"`, and `"grays"`.


```{r}
getColors("blues")
```

```{r echo=FALSE}
getColors("blues", output=TRUE)
```


To create a divergent color palette, specify beginning and an ending color palettes, which provide values for the  parameters `pal` and `end_pal`, where `pal` abbreviates palette. Here, generate colors from rust to blue.

```{r}
getColors("rusts", "blues")


```{r echo=FALSE}
getColors("rusts", "blues", output=TRUE)
```

## Utilities

[examples of utility functions](https://web.pdx.edu/~gerbing/lessR/examples/utilities.html)

**lessR** provides several utility functions for recoding, reshaping, and rescaling data.
