---
title: "Regression"
author: "David Gerbing"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{3. Models: Least-Squares, Logistic}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r}
library("lessR")
```

The `Regression()` function performs multiple facets of a complete regression analysis. Abbreviate with `reg()`.


To illustrate, first read the Employee data included as part of __lessR__. Read into the default __lessR__ data frame _d_.

```{r read}
d <- Read("Employee")
```


## Default Analysis 

### Brief output

The brief version provides just the basic analysis, what Excel provides, plus a scatterplot with the regression line, which becomes a scatterplot matrix with multiple regression.

```{r fig.width=4.5, fig.height=4.5}
reg_brief(Salary ~ Years + Pre)
```


### Full output

The full output is extensive: Summary of the analysis, estimated model, fit indices, ANOVA, correlation matrix, collinearity analysis, best subset regression, residuals and influence statistics, and prediction intervals. The idea is to provide all of the information you need for a proper regression analysis.

```{r}
reg(Salary ~ Years + Pre)
```

Brief output with standardization of all variables in the model. Plot the residuals as a line connecting each data point to the corresponding point on the regression line.

```{r, fig.width=4.5, fig.height=4}
reg_brief(Salary ~ Years, rescale="z", plot_errors=TRUE)
```


### k-fold cross-validation

Specify a cross-validation with the `kfold` parameter. Here specify three folds. The funciton automatically creates the training and testing data sets.

```{r}
reg(Salary ~ Years, kfold=3)
```

The standard output also includes $R^2_{press}$, the value of $R^2$ when applied to new, previously unseen data, a value comparable to the average $R^2$ on test data. 


## Output as a Stored Object

The output of `Regression()` can be stored into an R object, here named _r_. The output object consists of various components that together define a comprehensive regression analysis. R calls the resulting output structure a _list_ object.

```{r}
r <- reg(Salary ~ Years + Pre)
```

Entering the name of the object displays the full output.

```{r}
r
```

Or, work with the components individually. Use the base R `names()` function to identify all of the components. Component names that begin with `out_` are part of the standard output. Other components include just data and statistics designed to be input in additional procedures, including R markdown documents.

```{r}
names(r)
```

Here just display the estimates as part of the standard text output.

```{r}
r$out_estimates
```


Here display the coefficients as numeric values.

```{r}
r$coefficients
```

Do a regression and request all prediction intervals. Then convert that output segment to a data frame named _dp_ with base R `read.table()`. As a data frame, do a standard search for an individual row for a specific prediction interval (see the Subset a Data Frame vignette). 

This particular conversion to a data frame requires one more step. One more spaces delimit adjacent columns, but the names in this data set are formatted with a comma followed by a space. Use base R `sub()` to remove the space after the comma before converting to a data frame.

```{r}
r <- reg(Salary ~ Years, pred_rows="all", graphics=FALSE)
r$out_predict = sub(", ", ",", r$out_predict, fixed=TRUE)
dp <- read.table(text=r$out_predict)
dp[.(row.names(dp) == "Pham,Scott"),]
```


## Contrasts

Because `reg()` accomplishes its computations with base R function `lm()`, can pass `lm()` parameters to `reg()`, which then passes their values to `lm()`. Here first use base R function `contr.sum()` to calculate an effect coding contrast matrix for a categorical variable with three levels, such as the variable _Plan_ in the _Employee_ data set.

```{r}
cnt <- contr.sum(n=3)
cnt
```

Now use the `lm()` parameter `contrasts` to define the effect coding for _Plan_, passed to `reg_brief()`. Contrasts only apply to factors, so first convert _Plan_ to a factor.

```{r fig.width=4.5, fig.height=4}
d$Plan <- factor(d$Plan)
reg_brief(Salary ~ Plan, contrasts=list(Plan=cnt))
```


## Null Model

The $R^2$ fit statistic compares the sum of the squared errors of the model with the X predictor variables to the sum of squared errors of the null model.  The baseline of comparison, the null model, is a model with no X variables such that the fitted value for each set of X values is the mean of response variable $y$. The corresponding slope intercept is the mean of $y$, and the standard deviation of the residuals is the standard deviation of $y$.

The following submits the null model for _Salary_, and plots the errors. Compare the standard deviation of the residuals to a regression model of _Salary_ with one or more predictor variables.

```{r fig.width=5}
reg_brief(Salary ~ 1, plot_errors=TRUE)
```

Can also get the null model output from the __lessR__ function `Plot()` with the `fit` parameter set to `"null"`.


## Likert Type Data

When both variables consists of less than 10 unique integer values, the scatterplot is displayed as a bubble plot. This way there is no overprinting of the same point so that the number of values that represent a point is displayed.

```{r likert, fig.width=4.5, fig.height=4.5}
dd <- Read("Mach4")
reg_brief(m10 ~ m02, data=dd)
```



## Interpreted Output

The parameter `Rmd` creates an R markdown file that is automatically generated and then the corresponding html document from knitting the various output components together with full interpretation. A new, much more complete form of computer output.

Not run here.

```
reg(Salary ~ Years + Pre, Rmd="eg")
```

## Logistic Regression

Specify multiple logistic regression with the usual R formula syntax applied to the __lessR__ function `Logit()`. The output includes the confusion matrix and various classification fit indices. 


### Default Analysis

```{r fig.width=4, fig.height=4}
Logit(Gender ~ Salary)
```


### Change Classifiation Threshold

Specify additional probability thresholds for classification beyond just the default 0.5 with the `prob_cut` parameter.

```{r}
Logit(Gender ~ Years + Salary, prob_cut=c(.3, .5, .7))
```


## Full Manual

Use the base R `help()` function to view the full manual for `Regression()`. Simply enter a question mark followed by the name of the function, or its abbreviation.

```
?reg
```

