---
title: "Color Palettes"
author: "David Gerbing"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{color palettes}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r}
library("lessR")
```

The `getColors()` function can generate a variety of color palates. Use them to, for example, fill the bars in a histogram or bar chart with the  `fill` parameter.



## Qualitative Scales

The default is a color spectrum of 12 hcl colors presented in the order in which they are assigned to discrete levels of a categorical variable.

```{r}
getColors()
```

Display the color spectrum of 12 hcl colors ordered by hue from 0 by intervals of 360/12 = 30 degrees.

```{r}
getColors(in_order=TRUE)
```

Get a color wheel of 36 ordered hues around the wheel.

```{r}
getColors(n=36, shape="wheel", border="off")
```


Obtain deep rich colors for HCL qualitative scale.

```{r}
getColors(c=90, l=45)
```



## Sequential Scales

Generate an hcl blue sequence with `c=60` and vary `l`.
```{r}
getColors("blues", labels=FALSE)
```


Vary chroma for a yellow hcl sequence.

```{r}
getColors("browns", c=c(20,90), l=60)
```



## Divergent Scales

Generate seven colors from rust to blue.

```{r}
getColors("rusts", "blues", n=7)
```

Add a custom value of chroma, `c`, to make less saturated.
```{r}
getColors("rusts", "blues", n=7, c=45)
```



## Manual Specification of Colors

Individually specify colors.

```{r}
getColors(c("black", "blue", "red"))
```

Generate a custom sequential range of colors.

```{r}
getColors(pal="aliceblue", end_pal="blue")
```

