---
title: "Scatter Plots"
author: "David Gerbing"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{scatter plots}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r}
library("lessR")
```


First read the Employee data included as part of __lessR__.

```{r read}
d <- Read("Employee")
```

__lessR__ provides many versions of a scatter plot with its `Plot()` function.


## Two Variables

The regular scatterplot.

```{r sp}
Plot(Years, Salary)
```


The enhanced scatterplot with parameter `enhance`. 

```{r spEnhance}
Plot(Years, Salary, enhance=TRUE)
```


Map variable _Pre_ to the points with the `size` parameter, a bubble plot.

```{r size}
Plot(Years, Salary, size=Pre)
```

Plot against levels of categorical variable _Gender_ with the `by` parameter.

```{r by}
Plot(Years, Salary, by=Gender)
```


Two categorical variables result in a bubble plot of their joint frequencies.

```{r}
Plot(Dept, Gender)
```


## Distribution of a Single Variable

The default plot for a single continuous variable includes not only the scatterplot, but also the violin plot and box plot, with outliers identified. Call this plot the VBS plot.

```{r}
Plot(Salary)
```


For a single categorical variable, get the corresponding bubble plot of frequencies.

```{r}
Plot(Dept)
```


## Cleveland Dot Plot

The Cleveland dot plot, here for a single variable, has row names on the y-axis. The default plots sorts by the value plotted.

```{r}
Plot(Salary, row_names)
```

The standard scatterplot version of a Cleveland dot plot.

```{r}
Plot(Salary, row_names, sort_yx="0", segments_y=FALSE)
```

This Cleveland dot plot has two x-variables, indicated as a standard R vector with the `c()` function. In this situation the two points on each row are connected with a line segment. By default the rows are sorted by distance between the successive points.

```{r}
Plot(c(Pre, Post), row_names)
```


